Well-quasi-orderings on word languages

### 2025-01-10 src/amalgam.tex
fix mismatch between unions of products and actual «power chains»

\begin{lemma}
    \label{dwclosed-infixes-wqo:lem}
    Let $L \subseteq \Sigma^*$ be a \kl{downwards closed} language for the
    \kl{infix relation} that is \kl{well-quasi-ordered}. Then, the following
    are equivalent:
    {\renewcommand{\theenumi}{\roman{enumi}}
     \renewcommand{\labelenumi}{(\theenumi)}
    \begin{enumerate}
        \item\label{dwci-reg:item} $L$ is a \kl{regular language},
        \item\label{dwci-aml:item} $L$ is recognized by \emph{some} \kl{amalgamation system},
        \item\label{dwci-bod:item} $L$ is a \kl{bounded language},
        \item\label{dwci-uoc:item} $L$ is included in a finite union of products of \kl{chains}
    \end{enumerate}
    }
\end{lemma}
\begin{proof}
    It is clear that \cref{dwci-reg:item} $\Rightarrow$ \cref{dwci-aml:item}
    because regular languages are recognized by finite automata, and finite
    automata are a particular case of \kl{amalgamation systems}.
    The implication \cref{dwci-aml:item} $\Rightarrow$ \cref{dwci-bod:item}
    is the content of \cref{infix-amalgamation:thm}.
    The implication \cref{dwci-bod:item} $\Rightarrow$ \cref{dwci-uoc:item}
    is \cref{bounded-language:thm}.
    Finally, the implication \cref{dwci-uoc:item} $\Rightarrow$ \cref{dwci-reg:item}
    is simply because a \kl{downwards closed} language 
    that is a finite union of products of \kl{chains} is a regular language.

    \todo[inline]{This proof is not very satisfying.}
    Indeed, assume that
    $L$ is \kl{downwards closed} and included in a finite union of sets of the form
    $\InfPeriodChain{x} u \InfPeriodChain{y}$ where $x,y,u$ are possibly empty words.
    We can assume without loss of generality that
    for every $n$, $x^n u y^n$ is in $L$, otherwise, we have a bound on the maximal $n$ such that
    $x^n u y^n$ is in $L$, and we can increase the number of languages in the union, replacing $x$ or $y$
    with the empty word as necessary.
    Let us write $L' \defined \bigcup_{i = 1}^k x_i^* u_i y_i^*$. Then, $L'
    \subseteq L$ by construction. Furthermore, $L \subseteq \dwset{L'}$, also
    by construction. Finally, we conclude that $L = \dwset{L'}$ because $L$ is
    \kl{downwards closed}. Now, because $L'$ is a \kl{regular language}, and 
    \kl{regular languges} are closed under \kl{downwards closure}, we conclude
    that $L$ is a \kl{regular language}.
\end{proof}

>>>>

\begin{lemma}
    \label{dwclosed-infixes-wqo:lem}
    Let $L \subseteq \Sigma^*$ be a \kl{downwards closed} language for the
    \kl{infix relation} that is \kl{well-quasi-ordered}. Then, the following
    are equivalent:
    {\renewcommand{\theenumi}{\roman{enumi}}
     \renewcommand{\labelenumi}{(\theenumi)}
    \begin{enumerate}
        \item\label{dwci-reg:item} $L$ is a \kl{regular language},
        \item\label{dwci-aml:item} $L$ is recognized by \emph{some} \kl{amalgamation system},
        \item\label{dwci-bod:item} $L$ is a \kl{bounded language},
        \item\label{dwci-uoc:item} There exists 
            a finite set $E \subseteq (\Sigma^*)^3$
            such that $L = \bigcup_{(x,u,y) \in E} \InfPeriodChain{x} u \InfPeriodChain{y}$.
    \end{enumerate}
    }
\end{lemma}
\begin{proof}
    It is clear that \cref{dwci-reg:item} $\Rightarrow$ \cref{dwci-aml:item}
    because regular languages are recognized by finite automata, and finite
    automata are a particular case of \kl{amalgamation systems}.
    The implication \cref{dwci-aml:item} $\Rightarrow$ \cref{dwci-bod:item}
    is the content of \cref{infix-amalgamation:thm}.
    The implication \cref{dwci-bod:item} $\Rightarrow$ \cref{dwci-uoc:item}
    is \cref{bounded-language:lem}.
    Finally, the implication \cref{dwci-uoc:item} $\Rightarrow$ \cref{dwci-reg:item}
    is simply because a \kl{downwards closed} language 
    that is a finite union of products of \kl{chains} is a regular language.

    Indeed, assume that
    $L$ is \kl{downwards closed} and included in a finite union of sets of the form
    $\InfPeriodChain{x} u \InfPeriodChain{y}$ where $x,y,u$ are possibly empty words.
    We can assume without loss of generality that
    for every $n$, $x^n u y^n$ is in $L$, otherwise, we have a bound on the maximal $n$ such that
    $x^n u y^n$ is in $L$, and we can increase the number of languages in the union, replacing $x$ or $y$
    with the empty word as necessary.
    Let us write $L' \defined \bigcup_{i = 1}^k x_i^* u_i y_i^*$. Then, $L'
    \subseteq L$ by construction. Furthermore, $L \subseteq \dwset{L'}$, also
    by construction. Finally, we conclude that $L = \dwset{L'}$ because $L$ is
    \kl{downwards closed}. Now, because $L'$ is a \kl{regular language}, and 
    \kl{regular languges} are closed under \kl{downwards closure}, we conclude
    that $L$ is a \kl{regular language}.
\end{proof}

### 2025-01-10 src/infixes.tex
adding a lot of todos before arxiv
fix mismatch between unions of products and actual «power chains»

\begin{proofof}{bounded-language:thm}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

@@

    Now, admissible languages are ready to be pumped according to
    \cref{pumping-periods:lem}. For every admissible language,
    the size of a word that is not iterated is at most
    $N_0 \times m$ by definition, and we conclude that:
    \begin{equation}
        \label{bounded-language:eq}
        L \subseteq 
        \bigcup_{x,y \in \Sigma^{\leq n}}
        \bigcup_{u \in \Sigma^{\leq m \times N_0}}
        \InfPeriodChain{x} u \InfPeriodChain{y}
        \cup
        \InfPeriodChain{x} u
        \cup
        u \InfPeriodChain{x}
        \quad .
    \end{equation}
\end{proofof}

>>>>

\begin{lemma}
    \label{bounded-language:lem}
    Let $L \subseteq \Sigma^*$ be a \kl{bounded language}
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, there exists a finite subset $E \subseteq (\Sigma^*)^3$,
    such that:
    \begin{equation*}
        L \subseteq \bigcup_{(x,u,y) \in E} \InfPeriodChain{x} u \InfPeriodChain{y}
        \quad .
    \end{equation*}
\end{lemma}
\begin{proof}
    Let $w_1, \dots, w_n$ be such that
    $L \subseteq w_1^* \cdots w_n^*$.
    Let us define $m \defined \max \setof{\card{w_i}}{1 \leq i \leq n}$

@@

    Now, admissible languages are ready to be pumped according to
    \cref{pumping-periods:lem}. For every admissible language,
    the size of a word that is not iterated is at most
    $N_0 \times m$ by definition, and we conclude that:
    \begin{equation}
        \label{bounded-language:eq}
        L \subseteq 
        \bigcup_{x,y \in \Sigma^{\leq n}}
        \bigcup_{u \in \Sigma^{\leq m \times N_0}}
        \InfPeriodChain{x} u \InfPeriodChain{y}
        \cup
        \InfPeriodChain{x} u
        \cup
        u \InfPeriodChain{x}
        \quad .
    \end{equation}
\end{proof}

\begin{proofof}{bounded-language:thm}
    We apply \cref{bounded-language:lem}, and conclude
    because $\InfPeriodChain{x}$ is a finite union of \kl{chains}
    for the \kl{prefix}, \kl{suffix} and \kl{infix} relations
    (\cref{inf-period-chain:lem}).
\end{proofof}
